---
output:
  word_document: default
  html_document: default
---
```{r}
###############
##    1.1    ##
###############


set.seed(1234567890)
library(geosphere)

gaussian_kernel <- function(distance, h){
  return (exp(-(distance/h)^2))
}

#https://en.wikipedia.org/wiki/Geographical_center_of_Sweden
lat <- 62.173276 
long <- 14.942265

stations <- read.csv(file="stations.csv", fileEncoding="latin1", stringsAsFactors=FALSE)
stations <- cbind(stations,distHaversine(p1=stations[,c("latitude","longitude")],p2= c(lat,long)))
colnames(stations)[ncol(stations)] <- "distance"

temps <- read.csv(file="temps50k.csv", stringsAsFactors=FALSE)

st <- merge(stations,temps,by="station_number")

edge_stations <- vector(length=4)
edge_stations[1] <- stations[which.max(stations$latitude),"station_number"]
edge_stations[2] <- stations[which.min(stations$latitude),"station_number"]
edge_stations[3] <- stations[which.max(stations$longitude),"station_number"]
edge_stations[4] <- stations[which.min(stations$longitude),"station_number"]

threshold_station <- round(nrow(stations)/2)
max_distance <- 0

threshold_distances <- vector(length = length(edge_stations))

for (i in 1:length(edge_stations)){
   distances <- vector(length = nrow(stations))
   for (j in 1:nrow(stations)){
   distances[j] <- distHaversine(stations[which(stations$station_number==edge_stations[i]),c("longitude","latitude")], stations[j,c("longitude","latitude")])
   }
   distances <- sort(distances,decreasing = FALSE)
   threshold_distances[i] <- distances[threshold_station]
   max_distance <- round(ifelse(max_distance<max(distances),max(distances), max_distance))
}
threshold_distance <- mean(threshold_distances)

h_distance <- 820000

plot(x = seq(0,max_distance,length.out = 1000), y = gaussian_kernel(distance = seq(0,max_distance,length.out = 1000), h = h_distance), ylab = "Weight", xlab = "distance [m]")
points(x=threshold_distance, y = gaussian_kernel(distance = threshold_distance,h_distance), col="red", type = "p", cex=1.5)
text(x=threshold_distance*1.5, y = gaussian_kernel(distance = threshold_distance,h_distance)*1.5, col="red",labels= paste("Average distance to", threshold_station,"th closest station"))
points(x=threshold_distance, y = 0.5, col="green", type = "p", cex=2)
text(x=threshold_distance/2, y = 0.5, col="green", labels = "Weight=0.5")


h_date <- 107

plot(x=seq(0,180,length.out = 1000), y = gaussian_kernel(seq(0,180,length.out = 1000), h_date), ylab = "Weight", xlab = "Day difference", xlim=c(0,180), ylim=c(0,1))
points(x=90, y = 0.5, col="green", type = "p", cex=2)
text(x=60, y = 0.5, col="green", labels = "Weight = 0.5")
points(x=90, y = gaussian_kernel(90, h_date), col="red", type = "p", cex=1.5)
text(x=130, y = gaussian_kernel(90, h_date), col="red", labels = "90 days difference")


h_time <- 14.5

plot(x=seq(0,24,length.out = 1000), y = gaussian_kernel(seq(0,24,length.out = 1000), h_time), ylab = "Weight", xlab = "Hour difference", xlim=c(0,24), ylim=c(0,1))
points(x=12, y = 0.5, col="green", type = "p", cex=2)
text(x=8, y = 0.5, col="green", labels = "Weight = 0.5")
points(x=12, y = gaussian_kernel(12, h_time), col="red", type = "p", cex=1.5)
text(x=18, y = gaussian_kernel(12, h_time), col="red", labels = "12 hours difference")

```
The width of kernels were chosen so that the weight had decreased to 0.5 at a certain distance I figured was reasonable.

```{r}

date <- "2015-12-24" # The date to predict (up to the students)

times <- c("04:00:00", "06:00:00", "08:00:00","10:00:00","12:00:00","14:00:00","16:00:00","18:00:00","20:00:00","22:00:00", "24:00:00")

temp1 <- vector(length=length(times))
temp2 <- vector(length=length(times))

# Students’ code here

st <- st[which(st$date<=date),]

for (t in 1:length(times))
{
   exclude <- which((st$date==date)&(st$time>=times[t]))
   if (length(exclude)>0) {
      data <- st[-exclude,]
   } else {
      data <- st
   }
   
   rownames(data) <- 1:nrow(data)
   
   my_matrix <- matrix(NA,nrow=nrow(data), ncol=4)
   colnames(my_matrix) <- c("temperature", "k_sum", "k_prod", "index")
   
   for (s in 1:nrow(data)){
      k1 <- gaussian_kernel(data[s,"distance"], h_distance)

      days_diff <- as.numeric(as.Date(date)-as.Date(data[s,"date"]))%%365.25
      days_diff <- min(365-days_diff, days_diff)
      k2 <- gaussian_kernel(days_diff, h_date)

      hour_diff <- as.numeric(strptime(times[t],"%H:%M:%S")-strptime(data[s,"time"],"%H:%M:%S"))%%24
      hour_diff <- min(24-hour_diff, hour_diff)
      k3 <- gaussian_kernel(hour_diff,h_time)
      my_matrix[s,] <- c(data[s,"air_temperature"], k1+k2+k3, k1*k2*k3, s)
   }
   
   my_matrix[,"k_sum"] <- my_matrix[,"k_sum"]/sum(my_matrix[,"k_sum"])
   my_matrix[,"k_prod"] <- my_matrix[,"k_prod"]/sum(my_matrix[,"k_prod"])

   prediciton <- my_matrix[,"temperature"]*my_matrix[,"k_sum"]
   temp1[t] <- sum(prediciton)
   prediciton <- my_matrix[,"temperature"]*my_matrix[,"k_prod"]
   temp2[t] <- sum(prediciton)

   my_matrix <- my_matrix[order(my_matrix[,"k_sum"], decreasing = TRUE),]
   
   if (t%%3 == 0) {
      
      cat("\n\nMost weighted predictions at ", times[t],"with summed kernels:\n")

      for (i in 1:3)
      {
         print(unlist(data[my_matrix[i,"index"],c("distance","date", "time")]))
      }

      my_matrix <- my_matrix[order(my_matrix[,"k_prod"], decreasing = TRUE),]

      cat("\n\nMost weighted predictions at ", times[t],"with multiplied kernels:\n")

      for (i in 1:3)
      {
         print(unlist(data[my_matrix[i,"index"],c("distance","date", "time")]))
      }
   }
}
```


```{r}
ylim = c(min(min(temp1),min(temp2)), max(max(temp1),max(temp2)))
plot(temp1, type="o", ylim=ylim, col = "blue", ylab="Temperature prediction")
legend("left", legend=c("Summed kernels", "Multiplied kernels"),col=c("blue", "green"), lty = c(1,1), pch = c(1,1))
lines(temp2, type="o", col = "green")
```

The contribtution from each temperature measurement with the summed kernel gets a higher average weight, compared to the multiplied. This is because if any of the kernels generate weight near zero or equal to zero, the entire weight for that measurement becomes zero with the multiplied kernel

```{r}
RNGversion('3.5.1')
library(neuralnet)
par(mfrow=c(1,1))


set.seed(1234567890)
Var <- runif(50, 0, 10)
trva <- data.frame(Var, Sin=sin(Var))
tr <- trva[1:25,] 
va <- trva[26:50,] 
winit <- runif(31,-1,1)
MSE <- vector(length = 10)
ABS <- vector(length = 10)


for(i in 1:10) {
  nn <- neuralnet(Sin~Var, data=tr, hidden = 10, threshold = i/1000, startweights = winit)
  prediction <- compute(nn, as.matrix(va$Var))$net.result
  MSE[i] <- mean((prediction - va$Sin)^2)
  ABS[i] <- mean(abs(prediction - va$Sin))
}

xmin <- which.min(MSE)
ymin <- MSE[xmin]
plot(MSE, type = 'b', xlab = "Threshold*1000")
for (i in 1:length(MSE)){
   text(x = i, y = MSE[i], labels = round(MSE[i],4))

}
text(x=xmin, y = ymin, labels="MIN", col = "green")


nn <- neuralnet(Sin~Var, data=tr, hidden = 10, threshold = which.min(MSE)/1000, startweights = winit)
plot(nn)

# Plot of the predictions (black dots) and the data (red dots)
plot(trva, col = "red", cex = 0.8)
points(prediction(nn)$rep1)
```

